---
title: "Base R's built-in logit and expit functions"
description: |
  `plogis()` and `qlogis()`
author:
  - name: Quang Nguyen
    url: https://github.com/qntkhvn
date: "July 21, 2022"
output:
  distill::distill_article:
    highlight: rstudio
    code_folding: false
    self_contained: false
    toc: true
    toc_depth: 4
categories:
  - functions
---

I did not know there were built-in functions in base `R` for calculating the logit and expit... until today.

First, recall that the **logit** function for any $x \in (0,1)$ is

$$
logit(x) = log\left(\frac{x}{1-x}\right). 
$$
```{r, fig.height=5, echo = FALSE}
plot(\(x) log(x/(1-x)), ylab = "logit(x)")
```

When working with categorical data, we can think of $x$ here as a probability of success. Hence the logit of $x$ is the log odds ratio of the probability of success over the probability of failure.

To obtain the log odds for a given probability in `R`, we could do...

```{r}
prob <- 0.2
log(prob / (1 - prob))
```

or simply with `qlogis()`, as I just discovered...

```{r}
qlogis(prob)
```

On the other hand, the **expit** function for any $x \in \mathbb{R}$ is

$$
expit(x) = \frac{e^x}{1+e^x}.
$$
```{r, fig.height=5, echo = FALSE}
plot(\(x) exp(x)/(1+exp(x)), ylab = "expit(x)", xlim = c(-5, 5))
```

The expit maps the real numbers to $(0, 1)$. It is also known as the inverse logit function, or the sigmoid function, which is very popular in deep nets.

You've probably done this type of conversion before, in a logistic regression setting. When making a prediction after fitting a model and you're being ask to give a probability, this is exactly what you would do. 

Just as before, we could do it "by hand"...

```{r}
logit <- 0.3
exp(logit) / (1 + exp(logit))
```

or with a more simple function, `plogis()`...

```{r}
plogis(logit)
```

From my personal experience, I have written down the full formula for both logit and expit many times and forgotten a parenthesis here and there every so often. This is where `plogis` and `qlogis` will come in handy. More simple. Less code writing. And no more errors related to ()'s.
