---
title: Heatmap of the first 100000 digits of pi and e.
description: Chaos of numbers
author:
  - name: Quang Nguyen
    url: https://github.com/qntkhvn
date: "August 1, 2022"
output:
  distill::distill_article:
    highlight: rstudio
    code_folding: false
    self_contained: false
    toc: true
    toc_depth: 4
categories:
  - data viz
  - ggplot2
  - numbers
---

I recently saw this Twitter post about a heatmap of the first 100000 digits of $\pi$:

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">A visual representation of the first 100,000 digits of π <a href="https://t.co/MPGpckbds0">pic.twitter.com/MPGpckbds0</a></p>&mdash; Fermat&#39;s Library (@fermatslibrary) <a href="https://twitter.com/fermatslibrary/status/1553738762873602048?ref_src=twsrc%5Etfw">July 31, 2022</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

Let's see how closely I can recreate this figure.

First, let's get data for the first 100000 digits of $\pi$, clean it up, and transform it to the right format. I will let the figure size be 400 x 250.

```{r}
library(tidyverse)
library(rvest)

pi <- read_html("https://www.angio.net/pi/digits/100000.txt") |> 
  html_text() |> 
  str_remove("\\.") 

pi_digit <- tibble(digit = pi) |> 
  mutate(digit = str_split(digit, "")) |> 
  unnest_longer(digit) |> 
  mutate(row_index = rep(1:400, 250),
         col_index = rep(1:250, each = 400))
```

Now, a few lines of `ggplot2` should give the desired visualization.

```{r}
pi_digit |>
  ggplot(aes(row_index, col_index, fill = factor(digit))) +
  geom_tile() +
  scale_fill_manual(values = rainbow(10)) +
  scale_y_reverse() + 
  guides(fill = guide_legend(nrow = 1, title = NULL)) + 
  coord_equal() +
  theme_void() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal",
        legend.text = element_text(size = 9, margin = margin(l = -17, unit = "pt")))
```

Breakdown:

*   `geom_tile()` did the trick here. This one of my favorite `ggplot2` functions.

*   I recognized the color scheme in the original plot is quite similar to a rainbow palette. The function `rainbow()` came in handy here.

*   I also reversed the y-axis scale, to ensure the first digit (3) is on the top left. 

*   With a bit of digging, I was able to figure out how to move the legend text inside the colored keys.

Bonus: let's also plot the 100000 digits of $e$ (Euler's number).

```{r}
e <- read_html("http://boston.conman.org/2004/11/12/e100k.txt") |> 
  html_text() |> 
  str_remove("^[^\n]*\n") |> 
  str_remove_all("\\s")

e_digit <- tibble(digit = str_c(2, str_sub(e, 1, 99999))) |> 
  mutate(digit = str_split(digit, "")) |> 
  unnest_longer(digit) |> 
  mutate(row_index = rep(1:400, each = 250),
         col_index = rep(1:250, 400))

e_digit |>
  ggplot(aes(row_index, col_index, fill = factor(digit))) +
  geom_tile() +
  scale_fill_manual(values = rainbow(10)) +
  scale_y_reverse() + 
  guides(fill = guide_legend(nrow = 1, title = NULL))+ 
  coord_equal() +
  theme_void() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal",
        legend.text = element_text(size = 9, margin = margin(l = -17, unit = "pt")))
```

Both figures look noisy. There's no pattern at all for these two famous numbers.