---
title: "March Madness Seedigami"
description: |
  How does each seed perform?
author:
  - name: Quang Nguyen
    url: https://github.com/qntkhvn
date: "April 1, 2024"
output:
  distill::distill_article:
    highlight: rstudio
    code_folding: true
    self_contained: false
categories:
  - data viz
  - ggplot2
  - heatmap
preview: seedigami.png
---

March Madness is currently happening. Last night, I stumbled upon this tweet:

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Seedigami! NC State is the first 11 seed to beat a 4 seed in NCAA tournament history <a href="https://t.co/FbKcQ7Qwdi">pic.twitter.com/FbKcQ7Qwdi</a></p>&mdash; SaintArkweather (@SwordfishCFH) <a href="https://twitter.com/SwordfishCFH/status/1774581526371389448?ref_src=twsrc%5Etfw">March 31, 2024</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

In case you're not familiar, [Scorigami](https://en.wikipedia.org/wiki/Scorigami) is a fun concept in sports which refers to a game/match that finishes in a score that has not happened before in the history of a sport league.
This can be applied to any sport, but perhaps it is most popular in American football, and there's even a [Twitter account](https://twitter.com/nfl_scorigami) devoted to NFL Scorigami.

The tweet above is a nice little spin-off of Scorigami. It is Scorigami, but for March Madness seeding, hence the name "Seedigami". The purpose is to illustrate whether one seed has recorded a victory against other seeds in the history of the tournament. Last night, it was the first instance ever where an 11 seed (NC State) won a game against a 4 seed (Duke).

As soon as I saw this tweet, I immediately thought about making a slightly different version of it.
That is, a heatmap showing how often each seed has beaten every other seed throughout the history of March Madness.

Here, let's only consider every year starting from 1985, since the tournament was expanded to 64 teams that year.
I can easily get the data from the famous [March Machine Learning Mania Kaggle competition](https://www.kaggle.com/competitions/march-machine-learning-mania-2024/data), whose 10th edition is in progress.

What I'll need are the following tables (more detailed descriptions can be found in the data link above):

* `MNCAATourneyCompactResults`: for game results

* `MNCAATourneySeeds`: for seeding info

Now, I'll (briefly) walk through my entire process of creating the desired heat map, starting with data preparation. First, let's import the data and get a glimpse of both tables.

```{r}
library(tidyverse)
library(janitor)
results <- read_csv("MNCAATourneyCompactResults.csv") |> 
  clean_names()
seeds <- read_csv("MNCAATourneySeeds.csv") |> 
  clean_names()
```

```{r}
results
```

```{r}
seeds
```


Each tournament consists of 63 games. Notice that the data also include games from the [First Four](https://en.wikipedia.org/wiki/First_Four).

```{r}
results |> 
  count(season) |> 
  arrange(- n)
```

And TIL that the First Four used to be the "Opening round" (2001-2010), before the name was switched to First Four in 2011 and remains the same until now. These facts can be used in the data preparation process to keep only the 64 teams in the first round of every tournament.

```{r}
# main 64-team field
results_main <- results |> 
  group_by(season) |>
  mutate(game_num = row_number()) |> # define game number for each tournament
  ungroup() |>
  filter(!(season %in% 2001:2010 & game_num == 1)) |> 
  filter(!(season %in% 2011:2023 & game_num %in% 1:4))

results_main |> 
  count(season) |> 
  arrange(n)
```

Notice in 2021 there were only 62 games played instead of 63. I went back and check and found that there was a [no-contest game between VCU and Oregon in 2021](https://www.espn.com/mens-college-basketball/story/_/id/31102709/oregon-vcu-declared-no-contest-due-covid-19-protocols-ducks-advance) due to COVID, which explains why there was 1 less game. (Now I do remember this.)

What I want now is a table with the seeds for the winning and losing teams in the data. This can be done by pivoting the results table and then joining it with the seeding table.
Note that the ongoing 2024 tournament game results are not provided in the data, so I self-collected the 2024 results (as of today, up to the Elite 8).
(The 2024 data can probably be obtained via a freely-available data resource such as the [`hoopR`](https://hoopr.sportsdataverse.org/), [`ncaahoopR`](https://github.com/lbenz730/ncaahoopR), or [`cbbdata`](https://cbbdata.aweatherman.com) `R` packages, to name a few.)

```{r}
results_seeds <- results_main |> 
  select(season, w_team_id, l_team_id) |> 
  mutate(matchup = str_c(w_team_id, l_team_id)) |> 
  pivot_longer(ends_with("team_id"),
               names_to = "outcome",
               values_to = "team_id") |>
  mutate(outcome = str_remove(outcome, "_team_id")) |> 
  left_join(seeds)

results_seeds_w_l <- results_seeds |> 
  mutate(seed = str_remove_all(seed, "[:alpha:]"),
         seed = as.numeric(seed)) |> 
  select(-team_id) |> 
  pivot_wider(names_from = outcome,
              values_from = seed) |> 
  select(-matchup)

results24 <- tibble(
  season = 2024,
  w_l = c("1_16", "9_8", "5_12", "13_4", "11_6", "3_14", "7_10", "2_15",
          "1_9", "5_13", "3_11", "2_7", "1_5", "3_2", "1_3",
          "1_16", "9_8", "12_5", "4_13", "6_11", "3_14", "7_10", "2_15",
          "1_9", "4_12", "6_3", "2_7", "4_1", "6_2", "4_6",
          "1_16", "9_8", "12_5", "4_13", "11_6", "14_3", "10_7", "2_15",
          "1_9", "4_12", "11_14", "2_10", "4_1", "11_2", "11_4",
          "1_16", "8_9", "5_12", "4_13", "11_6", "3_14", "7_10", "2_15",
          "1_8", "5_4", "3_11", "2_7", "1_5", "2_3", "1_2")
) |> 
  separate(w_l, into = c("w", "l"), sep = "_", convert = TRUE)


results_seeds_w_l <- results_seeds_w_l |> 
  bind_rows(results24)
```

Now I have information on the winning and losing seeds for each game from 1985 up to today (April Fools' Day - but no fooling...). 

```{r}
results_seeds_w_l
```

How often has each of these combinations happened?

```{r}
results_seeds_w_l |> 
  count(w, l)
```

To handle the combinations that have never occurred, I create a table for every possible winning-losing seed combination, and then join with the existing table and update as needed.

```{r}
# initialize with all zero's
all_matchups <- crossing(w = 1:16, l = 1:16, n = 0)
# join and update
all_matchups <- all_matchups |> 
  rows_update(count(results_seeds_w_l, w, l), by = c("w", "l"))
all_matchups
```

Now we can finally make the heat map.

```{r, fig.height=7, fig.width=7, fig.align='center'}
all_matchups |> 
  ggplot(aes(w, l)) +
  geom_tile(aes(fill = n), color = "lightgray") +
  geom_text(aes(label = n, family = "Chivo"), size = rel(3)) +
  scale_x_continuous(breaks = 1:16) +
  scale_y_continuous(breaks = 1:16) +
  scale_fill_gradient2(high = "purple") +
  annotate("rect", 
           xmin = c(11, 15, 15, 3:5, 15, 16) - 0.5,
           xmax = c(11, 15, 15, 3:5, 15, 16) + 0.5,
           ymin = c(4, 3, 7, 3:5, 2, 1) - 0.5,
           ymax = c(4, 3, 7, 3:5, 2, 1) + 0.5, 
           color = "black", fill = "transparent", linewidth = 1) +
  labs(x = "\nWinning Seed",
       y = "Losing Seed\n",
       title = "March Madness Seedigami",
       subtitle = "How often does a seed beat another?") +
  theme_minimal(base_family = "Chivo") +
  theme(panel.grid = element_blank(),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

Some observations here:

* The highlighted cells are those that I find interesting.

* Teams don't beat opposing teams with the same seed number that frequently (a 3, 4, and 5 have beaten their corresponding similarly-seeded opponent only once).

* A 15 has beaten a 2 11 times. My favorite was Middle Tennessee beating Michigan State in 2016 (my first year following March Madness).

* More interesting facts about the 15 seed: They've also beaten a 3 seed once (Saint Peter's vs Purdue in 2022) and a 7 seed four times (which makes sense, given how often a 2 is beaten in the first round).

* As previously mentioned, NC State vs. Duke (11 vs. 4) yesterday was the latest Seedigami.

* And of course, the two famous 16 vs 1 upsets. (Will Purdue this year follow what Virginia did in 2019 - winning the title after being eliminated in the first round the year before?)

That's it for now... Maybe more March Madness stuff to come... MAYBE.

How many days have passed since my last blog?

```{r}
today() - as.Date("2022-12-19")
```

Until next time...



