---
title: LaTeX Font for Data Viz in R
description: |
  For future reference
author:
  - name: Quang Nguyen
    url: https://github.com/qntkhvn
date: "April 21, 2022"
output:
  distill::distill_article:
    highlight: rstudio
    code_folding: false
    self_contained: false
    toc: true
    toc_depth: 4
categories:
  - data viz
  - ggplot2
  - LaTeX
preview: latex-ggplot.png
---

So... I had been wondering about how to change the font when making data viz in R to the $\LaTeX$ font ([Computer Modern](https://www.overleaf.com/learn/latex/Font_typefaces)). Turns out, it's not that complicated.

###  Set up

(Note: The set up shown below is for a Mac device. I imagine it'd be something similar for Windows.)

*   I first downloaded the Computer Modern font (with `.ttf` file extension) from the internet. It is available on a few different websites, and the one I ended up with was https://www.fontsquirrel.com/fonts/computer-modern.

*   I then imported the unzipped folder of the downloaded font file to the Font Book app.

And... the set up is basically it. Now, let's make some data viz. First, let's quickly generate some data.

```{r}
set.seed(100)
x <- rnorm(100)
y <- x + rnorm(100)
```

I'm going to illustrate this for both `base` plotting and `ggplot2`.

### `base` `R`

By default, the font of a `base` `R` plot looks like this:

```{r, fig.height = 5.5, fig.width = 5.5}
plot(x,
     y,
     main = "A simple scatterplot",
     xlab = "Explanatory Variable",
     ylab = "Response Variable")
```

To apply the $\LaTeX$ font (or any font you want), use the good ol' `par()` function, and specify `family = "the font name"` inside of `par()`. Here, the name of the font is "CMU Serif".

```{r, fig.height = 5.5, fig.width = 5.5}
par(family = "CMU Serif")
plot(x,
     y,
     main = "A Simple Scatterplot",
     xlab = "Explanatory Variable",
     ylab = "Response Variable")
```

(I'm sure there are other ways of doing this in `base` `R` too...)

### `ggplot2`

For `ggplot2`, the default theme appears like this:

```{r, fig.height = 5, fig.width = 5}
library(tidyverse)
p <- tibble(x, y) %>% 
  ggplot(aes(x, y)) +
  geom_point() +
  labs(x = "Explanatory Variable",
       y = "Response Variable",
       title = "A Simple Scatterplot")
p
```

We can then call `theme()` to customize the text font. 

```{r, fig.height = 5, fig.width = 5}
p +
  theme(text = element_text(family = "CMU Serif"))
```

Of course, we can also do this separately for the different individual [theme components](https://ggplot2.tidyverse.org/reference/theme.html) (`axis.text`, `plot.title`, etc.) of `ggplot2`. 

Ta-da!

(I haven't blogged in awhile. This is my first blog since last August, before I moved to Chicago. Feels great to be back blogging again.)
